Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2018-08-16 sims/example/simple_example.Rmd
adjustments to simple tests

@@

```{r simple_data, cache=TRUE}
base_covmat <- matrix(c(1.0, 0.7, 0.0,
                        0.7, 1.0, 0.0,
                        0.0, 0.0, 1.0), nrow=3)
pop_covmats <- list(base_covmat, 
                    base_covmat[3:1, 3:1], 
                    base_covmat[c(1,3,2), c(1,3,2)])
struct_breakpoints <- c(0.0, 0.25, 0.5, 0.75, 1.0)
struct_indices <- c(1,2,2,3)
nsamples <- c(50, 50, 50)
ind_covmat <- diag(sum(nsamples))
for (k in seq_along(nsamples)) {
    ii <- sum(nsamples[seq_len(k-1)]) + 1:nsamples[k] 
    ind_covmat[ii, ii] <- rexp(nsamples[k]^2, 100)
    ind_covmat[ii, ii] <- ind_covmat[ii, ii] + t(ind_covmat[ii, ii])
}
pop_chols <- lapply(lapply(pop_covmats, chol),
                    function (x) { 
                        (x[rep(1:3, nsamples), rep(1:3, nsamples)] / sqrt(rep(nsamples, nsamples))
                         + ind_covmat)
                    })
nwindows <- 99
nloci <- 100
win.fn <- function (n, ...) {
    geno <- matrix(rnorm(sum(nsamples) * nloci), nrow=nloci)
    cholmat <- pop_chols[[struct_indices[findInterval(n/nwindows, struct_breakpoints, rightmost.closed=TRUE)]]]
    return( geno %*% cholmat )
}
attr(win.fn, "max.n") <- nwindows
attr(win.fn,"samples") <- paste0(rep(letters[seq_along(nsamples)], nsamples),
                                 unlist(lapply(nsamples, seq_len)))
attr(win.fn,"region") <- function (n) { 
    data.frame( chrom="matrix", start=((n-1)*nloci+1), end=n*nloci ) }
all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows)
regions <- region(win.fn)()
all.lengths <- c("chrom1"=nwindows)
pc.distmat <- pc_dist( all.pcas, npc=2 )
na.inds <- is.na( all.pcas[,1] ) # there may be windows with missing data
mds.coords <- cbind( data.frame( 
                        chrom=rep(names(all.lengths),all.lengths),
                        window=unlist(lapply(all.lengths,seq_len)) ),
                        cmdscale( pc.distmat[!na.inds,!na.inds], k=2 )[ ifelse( na.inds, NA, cumsum(!na.inds) ), ] )
mds.cols <- (1:ncol(mds.coords))[-(1:2)]
```

@@

Now we'll do the same thing, but with 50% missing data
on the first half of the chromosome.
```{r missing_data, cache=TRUE, depends="simple_data"}
win.fn <- function (n, ...) {
    geno <- matrix(rnorm(sum(nsamples) * nloci), nrow=nloci)
    cholmat <- pop_chols[[struct_indices[findInterval(n/nwindows, struct_breakpoints, rightmost.closed=TRUE)]]]
    out <- geno %*% cholmat
    if (n < nwindows/2) {
        out[rbinom(length(out), 1, 0.5)==1] <- NA
    }
    return(out)
}
attr(win.fn, "max.n") <- nwindows
attr(win.fn,"samples") <- paste0(rep(letters[seq_along(nsamples)], nsamples),
                                 unlist(lapply(nsamples, seq_len)))
attr(win.fn,"region") <- function (n) { 
    data.frame( chrom="matrix", start=((n-1)*nloci+1), end=n*nloci ) }
all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows)
pc.distmat <- pc_dist( all.pcas, npc=2 )
na.inds <- is.na( all.pcas[,1] ) # there may be windows with missing data
mds.coords <- cbind( data.frame( 
                        chrom=rep(names(all.lengths),all.lengths),
                        window=unlist(lapply(all.lengths,seq_len)) ),
                        cmdscale( pc.distmat[!na.inds,!na.inds], k=2 )[ ifelse( na.inds, NA, cumsum(!na.inds) ), ] )
mds.cols <- (1:ncol(mds.coords))[-(1:2)]

@@

Now we'll do the same thing, but with very different sample sizes.
```{r sample_diffs, cache=TRUE, depends="simple_data"}
nsamples <- c(900, 10, 10)
ind_covmat <- diag(sum(nsamples))
for (k in seq_along(nsamples)) {
    ii <- sum(nsamples[seq_len(k-1)]) + 1:nsamples[k] 
    ind_covmat[ii, ii] <- rexp(nsamples[k]^2, 15)
    ind_covmat[ii, ii] <- ind_covmat[ii, ii] + t(ind_covmat[ii, ii])
}
pop_chols <- lapply(lapply(pop_covmats, chol),
                    function (x) { 
                        (x[rep(1:3, nsamples), rep(1:3, nsamples)] / sqrt(rep(nsamples, nsamples))
                         + ind_covmat)
                    })
nwindows <- 99
nloci <- 50
win.fn <- function (n, ...) {
    geno <- matrix(rnorm(sum(nsamples) * nloci), nrow=nloci)
    cholmat <- pop_chols[[struct_indices[findInterval(n/nwindows, struct_breakpoints, rightmost.closed=TRUE)]]]
    return( geno %*% cholmat )
}
attr(win.fn, "max.n") <- nwindows
attr(win.fn,"samples") <- paste0(rep(letters[seq_along(nsamples)], nsamples),
                                 unlist(lapply(nsamples, seq_len)))
attr(win.fn,"region") <- function (n) { 
    data.frame( chrom="matrix", start=((n-1)*nloci+1), end=n*nloci ) }
all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows)
all.lengths <- c("chrom1"=nwindows)
pc.distmat <- pc_dist( all.pcas, npc=2 )
na.inds <- is.na( all.pcas[,1] ) # there may be windows with missing data
mds.coords <- cbind( data.frame( 
                        chrom=rep(names(all.lengths),all.lengths),
                        window=unlist(lapply(all.lengths,seq_len)) ),
                        cmdscale( pc.distmat[!na.inds,!na.inds], k=2 )[ ifelse( na.inds, NA, cumsum(!na.inds) ), ] )
mds.cols <- (1:ncol(mds.coords))[-(1:2)]

>>>>

**Note:** to copy the resulting figures to the `writeup/` directory:
```
for x in figure/simple_example/*pdf; do mv $x "../../writeup/simple_example_$(basename ${x%-1.pdf}.pdf)"; done
```

@@

```{r simple_data, cache=TRUE}
base_covmat <- matrix(c(0.75, 0.25, 0.0,
                        0.25, 0.75, 0.0,
                        0.0, 0.0, 1.0), nrow=3)
pop_covmats <- list(base_covmat, 
                    base_covmat[3:1, 3:1], 
                    base_covmat[c(1,3,2), c(1,3,2)])
struct_breakpoints <- c(0.0, 0.25, 0.5, 0.75, 1.0)
struct_indices <- c(1,2,2,3)
nsamples <- c(50, 50, 50)
# ind_covmat <- diag(sum(nsamples))
# for (k in seq_along(nsamples)) {
#     ii <- sum(nsamples[seq_len(k-1)]) + 1:nsamples[k] 
#     ind_covmat[ii, ii] <- rexp(nsamples[k]^2, 100)
#     ind_covmat[ii, ii] <- ind_covmat[ii, ii] + t(ind_covmat[ii, ii])
# }
ind_covmat <- 0
pop_chols <- lapply(lapply(pop_covmats, chol),
                    function (x) { 
                        (x[rep(1:3, nsamples), rep(1:3, nsamples)] / sqrt(rep(nsamples, nsamples))
                         + ind_covmat)
                    })
nwindows <- 99
nloci <- 400
win.fn <- function (n, ...) {
    geno <- matrix(rnorm(sum(nsamples) * nloci), nrow=nloci)
    cholmat <- pop_chols[[struct_indices[findInterval(n/nwindows, struct_breakpoints, rightmost.closed=TRUE)]]]
    return( geno %*% cholmat )
}
attr(win.fn, "max.n") <- nwindows
attr(win.fn,"samples") <- paste0(rep(letters[seq_along(nsamples)], nsamples),
                                 unlist(lapply(nsamples, seq_len)))
attr(win.fn,"region") <- function (n) { 
    data.frame( chrom="matrix", start=((n-1)*nloci+1), end=n*nloci ) }
all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows)
regions <- region(win.fn)()
all.lengths <- c("chrom1"=nwindows)
pc.distmat <- pc_dist( all.pcas, npc=2 )
na.inds <- is.na( all.pcas[,1] ) # there may be windows with missing data
mds.coords <- cbind( data.frame( 
                        chrom=rep(names(all.lengths),all.lengths),
                        window=unlist(lapply(all.lengths,seq_len)) ),
                        cmdscale( pc.distmat[!na.inds,!na.inds], k=2 )[ ifelse( na.inds, NA, cumsum(!na.inds) ), ] )
mds.cols <- (1:ncol(mds.coords))[-(1:2)]
```

@@

Now we'll do the same thing, but with 50% missing data
on the first half of the chromosome.
```{r missing_data, cache=TRUE, depends="simple_data"}
win.fn <- function (n, ...) {
    geno <- matrix(rnorm(sum(nsamples) * nloci), nrow=nloci)
    cholmat <- pop_chols[[struct_indices[findInterval(n/nwindows, struct_breakpoints, rightmost.closed=TRUE)]]]
    out <- geno %*% cholmat
    if (n < nwindows/2) {
        out[rbinom(length(out), 1, 0.5)==1] <- NA
    }
    return(out)
}
attr(win.fn, "max.n") <- nwindows
attr(win.fn,"samples") <- paste0(rep(letters[seq_along(nsamples)], nsamples),
                                 unlist(lapply(nsamples, seq_len)))
attr(win.fn,"region") <- function (n) { 
    data.frame( chrom="matrix", start=((n-1)*nloci+1), end=n*nloci ) }
all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows)
pc.distmat <- pc_dist( all.pcas, npc=2 )
na.inds <- is.na( all.pcas[,1] ) # there may be windows with missing data
mds.coords <- cbind( data.frame( 
                        chrom=rep(names(all.lengths),all.lengths),
                        window=unlist(lapply(all.lengths,seq_len)) ),
                        cmdscale( pc.distmat[!na.inds,!na.inds], k=2 )[ ifelse( na.inds, NA, cumsum(!na.inds) ), ] )
mds.cols <- (1:ncol(mds.coords))[-(1:2)]
plot_everything()
```

@@

Now we'll do the same thing, but with very different sample sizes.
```{r sample_diffs, cache=TRUE, depends="simple_data"}
nsamples <- c(900, 10, 10)
ind_covmat <- diag(sum(nsamples))
for (k in seq_along(nsamples)) {
    ii <- sum(nsamples[seq_len(k-1)]) + 1:nsamples[k] 
    ind_covmat[ii, ii] <- rexp(nsamples[k]^2, 15)
    ind_covmat[ii, ii] <- ind_covmat[ii, ii] + t(ind_covmat[ii, ii])
}
pop_chols <- lapply(lapply(pop_covmats, chol),
                    function (x) { 
                        (x[rep(1:3, nsamples), rep(1:3, nsamples)] / sqrt(rep(nsamples, nsamples))
                         + ind_covmat)
                    })
nwindows <- 99
nloci <- 400
win.fn <- function (n, ...) {
    geno <- matrix(rnorm(sum(nsamples) * nloci), nrow=nloci)
    cholmat <- pop_chols[[struct_indices[findInterval(n/nwindows, struct_breakpoints, rightmost.closed=TRUE)]]]
    return( geno %*% cholmat )
}
attr(win.fn, "max.n") <- nwindows
attr(win.fn,"samples") <- paste0(rep(letters[seq_along(nsamples)], nsamples),
                                 unlist(lapply(nsamples, seq_len)))
attr(win.fn,"region") <- function (n) { 
    data.frame( chrom="matrix", start=((n-1)*nloci+1), end=n*nloci ) }
all.pcas <- eigen_windows(win.fn, k=2, do.windows=1:nwindows)
all.lengths <- c("chrom1"=nwindows)
pc.distmat <- pc_dist( all.pcas, npc=2 )
na.inds <- is.na( all.pcas[,1] ) # there may be windows with missing data
mds.coords <- cbind( data.frame( 
                        chrom=rep(names(all.lengths),all.lengths),
                        window=unlist(lapply(all.lengths,seq_len)) ),
                        cmdscale( pc.distmat[!na.inds,!na.inds], k=2 )[ ifelse( na.inds, NA, cumsum(!na.inds) ), ] )
mds.cols <- (1:ncol(mds.coords))[-(1:2)]

### 2018-08-16 writeup/local_pca_paper.tex
adjustments to simple tests

@@

@@

>>>>

\paragraph{Testing}
Figure \ref{fig:test_missing} shows tests with lots of missing data.
Also the covariance matrix was chosen to have the top two eigenvalues equal,
so that eigenvector switching is expected.

@@

\paragraph{Testing}
To demonstrate the utilitiy of this method XXX
see Figure \ref{fig:test_weighted_pca}.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=0.82\textwidth]{simple_example_base_case}
        \includegraphics[width=0.82\textwidth]{simple_example_missing_data}
    \end{center}
    \caption{
        From simple example -- base case; and with 50\% missing data.
    } \label{fig:test_missing}.
\end{figure}

\begin{figure}
    \begin{center}
        \includegraphics[width=0.82\textwidth]{simple_example_sample_diffs}
        \includegraphics[width=0.82\textwidth]{simple_example_reweighted}
    \end{center}
    \caption{
        From simple example -- different sample sizes; unweighted (top); weighted by $1/\sqrt{n_i}$ (bottom).
    } \label{fig:test_weighted_pca}.
\end{figure}
